Mixture Models for Single-Cell Assays with Application to Vaccine Studies: Supplementary Materials

### 2013-05-17 MIMOSA.tex
Updated competing ROC in SI and varying number of observations in SI

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
 While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response.
 After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}. The antigen--specific T-cell subpopulations (i.e. the subset that can respond to one specific antigen) constitute a very small fraction of the total number of CD4 and CD8 T-cells. The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
Individual cells are labelled using fluorescently conjugated antibodies against cell--surface biomarkers (CD3, CD4, and CD8), used to subset T-cells, and functional biomarkers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected.
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.

These counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences. Subjects who generate a response after stimulation are called \textit{responders}, whereas subjects that do not show any differences are called \textit{non-responders}.
In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.

@@

\textit{Flow cytometry}:
The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen (prime at zero and two months, boost at four and six months)~\citep{Goepfert:2011ci}.
The goal of this data set was to assess the immune response to the vaccine across multiple antigen stimulations, time points, cytokines and T-cell subsets.
Here, we analyze a subset of the data consisting of 98 subjects from the vaccine group at two time points: day 0 and day 182.
Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured at the single-cell level for each subject and time point, with and without stimulations with an antigen (here we focus on HIV Envelope peptide pool) matching part of the vaccine. For ease of presentation we restricted ourselves to the CD4+ T-cell subsets. Samples on day 0 were taken just before vaccination and no response is expected there. The corresponding samples can be used as negative controls. Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least.

\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen subjects. T-cells isolated by flow cytometry from sixteen subjects were stimulated in blocks of four subjects with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, and CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays. The expression from the simulated samples  was compared to paired, unstimulated controls.

>>>>

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
 While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response.
 After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}. The antigen--specific T-cell subpopulations (i.e. the subset that can respond to one specific antigen) constitute a very small fraction of the total number of CD4 and CD8 T-cells. The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
Individual cells are labelled using fluorescently conjugated antibodies against cell--surface biomarkers (CD3, CD4, and CD8), used to subset T-cells, and functional biomarkers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that the rare cell populations can be detected.
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.

These counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences. Subjects who generate a response after stimulation are called \textit{responders}, whereas subjects that do not show any differences are called \textit{non-responders}.
In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (on the order of 10s of cells, with effect sizes on the order of $10^{-4}$, relative to the total number of cells), and real biological differences might be difficult to detect.

@@

\textit{Flow cytometry}:
The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen (prime at zero and two months, boost at four and six months)~\citep{Goepfert:2011ci}.
The goal of this data set was to assess the immune response to the vaccine across multiple antigen stimulations, time points, cytokines and T-cell subsets.
Here, we analyze a subset of the data consisting of 98 subjects from the vaccine group at two time points: day 0 and day 182.
Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured at the single-cell level for each subject and time point, with and without stimulations with an antigen (here we focus on HIV Envelope peptide pool) matching part of the vaccine. For ease of presentation we restricted ourselves to the CD4+ T-cell subsets. Samples on day 0 were taken just before vaccination and no response is expected there. The corresponding samples can be used as negative controls. Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least. In this data set a median of 51K and 58K T--cells were collected for the stimulated and unstimulated samples, respectively (IQR $\approx$ 37K and 43K, respectively) were collected for the stimulated and unstimulated samples. The effect sizes (differences in the proportions of stimulated and unstimulated cells) are on the order of $10^{-4}$, with the number of positive cells typically ranging from 1--70, with a few samples exhibiting large numbers of positive events.

\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen subjects. T-cells isolated by flow cytometry from sixteen subjects were stimulated in blocks of four subjects with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, and CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays. The expression from the simulated samples  was compared to paired, unstimulated controls. In this data set, we have approximately 90 single cells per subject per stimulation condition with 96 gene expression measurements per cell.

### 2013-05-17 SupplementaryMaterialMIMOSA.tex
Updated competing ROC in SI and varying number of observations in SI

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto,node distance=0cm]
    \node at (0,0) (A) {
    \begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.75	\columnwidth]{Figures/Sim_OneSided_ROC_varyNobs.pdf}};
         \begin{scope} [x={(c.south east)},y={(c.north west)}]
                 \node at (0,1) [font=\tiny\sffamily] {A} ;
	\end{scope}
     \end{tikzpicture}
     };
     \node [below=of A] (B) {
     \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.75\columnwidth]{Figures/Sim_OneSided_FDR_varyNobs.pdf}};
        \begin{scope} [x={(d.south east)},y={(d.north west)}]
            \node at (0,1) [font=\tiny\sffamily] {B} ;
\end{scope}
    \end{tikzpicture}
    };
  \end{tikzpicture}
   \caption{One--sided MIMOSA model fit to simulated data with 50K events and varying values of I (number of observations). A) Average ROC curves from 10 simulations for 20, 50 and 100 observations. B) Average observed and nominal FDR from 10 simulations for 20, 50, and 100 observations. This figure appears in color in the electronic version of this article.
}
   \label{webfig:simulations_Nobs}
\end{figure}
\clearpage

>>>>

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto,node distance=0cm]
    \node at (0,0) (A) {
    \begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.75	\columnwidth]{Figures/Sim_OneSided_ROC_varyNobs.pdf}};
         \begin{scope} [x={(c.south east)},y={(c.north west)}]
                 \node at (0,1) [font=\tiny\sffamily] {A} ;
	\end{scope}
     \end{tikzpicture}
     };
     \node [below=of A] (B) {
     \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.75\columnwidth]{Figures/Sim_OneSided_FDR_varyNobs.pdf}};
        \begin{scope} [x={(d.south east)},y={(d.north west)}]
            \node at (0,1) [font=\tiny\sffamily] {B} ;
\end{scope}
    \end{tikzpicture}
    };
  \end{tikzpicture}
   \caption{One--sided MIMOSA model fit to simulated data with 50K events and varying values of I (number of observations). A) Average ROC curves from 10 simulations for 20, 50 and 100 observations. B) Average observed and nominal FDR from 10 simulations for 20, 50, and 100 observations. This figure appears in color in the electronic version of this article.
}
   \label{webfig:simulations_Nobs}
\end{figure}
\clearpage
\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto,node distance=0cm]
    \node at (0,0) (A) {
    \begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.75	\columnwidth]{Figures/competeROC.pdf}};
         \begin{scope} [x={(c.south east)},y={(c.north west)}]
                 \node at (0,1) [font=\tiny\sffamily] {A} ;
	\end{scope}
     \end{tikzpicture}
     };
     \node [below=of A] (B) {
     \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.75\columnwidth]{Figures/competeROCfdr.pdf}};
        \begin{scope} [x={(d.south east)},y={(d.north west)}]
            \node at (0,1) [font=\tiny\sffamily] {B} ;
\end{scope}
    \end{tikzpicture}
    };
  \end{tikzpicture}
   \caption{Effect of deviations of the assumptions of 100\% true responders at the post-vaccine time point on A) ROC and B) FDR curves. The true response rate is shown in the gray box above each plot.
}
   \label{webfig:simulations_Nobs}
\end{figure}
\clearpage
